Modelling the emergence of whisker barrels

### 2020-07-11 paper/briefpaper/brief.tex
Makes changes suggested by Leah

The barrel boundaries form a Voronoi tessellation \citep{senft_mouse_1991}
(Fig.\,\ref{fig:main}A), suggesting that barreloid topology is preserved in
the projection of thalamocortical axons into the cortex, and that a barrel
forms by lateral axon branching from an initial center-point that ceases upon
contact with axons branching from adjacent centers.  However, the assumption
of pre-arranged center-points is difficult to resolve with the observation
that axons arrive in the cortical plate as an undifferentiated bundle,
\emph{prior} to barreloid formation \citep{agmon_organized_1993}.  \mpsix{In
  mice, axons from the trigeminal ganglion arrive in the principal sensory
  nucleus of the trigeminal nerve (PrV) at E12, then axons from the PrV arrive
  in the ventroposteromedial nucleus of the thalamus (VPM) at E17, then axons
  from the VPM arrive in the cortical plate at E18/P0. Distinct
  whisker-related clusters then become apparent in the PrV at P0-P1, in the
  VPM at P2-P3, and in the cortex at P3-P5}
\citep{erzurumlu_development_2012,sehara_neuronal_2011}.

@@

\mpfive{Testing this reasoning required increasing the strength of the
  competition between simulated thalamocortical projections for cortical
  territory, by increasing the tendency for each projection to compete for
  cortical space into which to concentrate. The major modification required
  was thus to introduce into the model an additional source of competition
  between thalamic projections.}
%
The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching,
\MPtwo{whilst ensuring that $a_i$ is conserved over time,} we
redefined
%
\begin{equation} \label{eq:comp}
  \color{colmptwo}
  \chi_i(\mb{x}, t) = \frac{\epsilon a_i}{N-1} \nabla \textstyle\sum_{j\ne i}^{N}{a_j}.
  \color{black}
\end{equation}
%
\mpfive{This term contributes to the \emph{flux} of axonal branching as an
  additional source of diffusion, scaled by $\epsilon$, which reduces the
  branching density for a given projection where the branches of other
  projections are dense. Note that this operation is local to individual
  afferent projections.}

\MPone{In addition, the model we have outlined requires that molecular
  guidance gradients in the cortex are complemented by graded values of the
  interaction strengths, $\gamma$, at the level of the thalamus. While the
  precise mechanisms by which thalamic and cortical gradients interact during
  development have not been fully characterised, the presence of complementary
  thalamic and cortical guidance gradients has been well established
  experimentally. In particular, the EphA4 receptor and its ligand ephrin-A5
  are distributed in complementary gradients in the somatosensory thalamus and
  cortex} (\citealp{vanderhaeghen_mapping_2000,miller_epha7-ephrin-a5_2006}).
\MPone{Cells originating in the medial part of ventrobasal thalamus
  (comprising VPM and VPL) express high levels of EphA receptors and project
  to the lateral part of S1, which expresses low levels of ephrin-A5, and
  cells originating in the lateral part of ventrobasal thalamus express low
  levels of EphA receptors and project to the medial part of S1, which
  expresses high levels of ephrin-A5} (see
\citealp{gao_regulation_1998,dufour_area_2003,vanderhaeghen_developmental_2004,speer_grading_2005,torii_role_2013}).
\MPone{We assume that such patterning arises because the relative strengths of
  interaction with guidance molecules (e.g., ephrin-A5) in the cortex are
  correlated with the relative concentrations of complementary molecules
  (e.g., EphA4) in the thalamus, and thus with thalamic position along the
  axis to which their gradients are aligned.}

@@

\MPthreePar{The parameters of the example simulation are indicated in
  Fig.\,\ref{fig:paramsweep}A using an asterisk. In
  Fig.\,\ref{fig:paramsweep}B, we also present examples of alternative
  patterns that emerge for different choices of $D$. Decreasing the rate of
  diffusion may be considered equivalent to increasing the overall size of the
  domain, $S$, hence insights into barrel development in species with
  substantially larger barrel fields, such as capybara which do not have
  clearly identifiable barrels, may be gained by studying pattern formation
  when $D$ is small. In this context, it is interesting to note that for small
  $D$, the organization is predicted to be topological but highly irregular,
  with a general expansion in the territory occupied by the central versus
  peripheral domains that would presumably manifest as an absence of
  identifiable barrel fields (Fig.\,\ref{fig:paramsweep}B\,iii).}

@@

\cmnt{In reality, scores of smaller Dirichletform barrels representing the
  microvibrissae form to the right of where the E-row barrels emerge in our
  simulations, presumably via the same competitive processes. Enforcing here
  the same boundary condition as used to represent the true edges of the
  barrel field was necessary to ensure the stability of the simulation, though
  we acknowledge that this region of the boundary was enforced primarily to
  keep the number of simulated projections, and hence the overall
  computational complexity of the simulation, manageable (simulating an extra
  projection introduces 13030 new dynamical variables).}

>>>>

The barrel boundaries form a Voronoi tessellation \citep{senft_mouse_1991}
(Fig.\,\ref{fig:main}A), suggesting that barreloid topology is preserved in
the projection of thalamocortical axons into the cortex, and that a barrel
forms by lateral axon branching from an initial center-point that ceases upon
contact with axons branching from adjacent centers.  However, the assumption
of pre-arranged center-points is difficult to resolve with the observation
that axons arrive in the cortical plate as an undifferentiated bundle,
\emph{prior} to barreloid formation \citep{agmon_organized_1993}.  \mpsix{In
  mice, axons from the trigeminal ganglion arrive in the principal division of
  the trigeminal nucleus (PrV) at E12, then axons from the PrV arrive
  in the ventroposteromedial nucleus of the thalamus (VPM) at E17, then axons
  from the VPM arrive in the cortical plate at E18/P0. Distinct
  whisker-related clusters then become apparent in the PrV at P0-P1, in the
  VPM at P2-P3, and in the cortex at P3-P5}
\citep{erzurumlu_development_2012,sehara_neuronal_2011}.

@@

\mpfive{Testing this reasoning required increasing the strength of the
  competition between simulated thalamocortical projections for cortical
  territory, by increasing the tendency for each projection to compete for
  cortical space in which to branch and make connections. The major
  modification required
  was thus to introduce into the model an additional source of competition
  between thalamic projections.}
%
The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching,
\MPtwo{whilst ensuring that $a_i$ is conserved over time,} we
redefined
%
\begin{equation} \label{eq:comp}
  \color{colmptwo}
  \chi_i(\mb{x}, t) = \frac{\epsilon a_i}{N-1} \nabla \textstyle\sum_{j\ne i}^{N}{a_j}.
  \color{black}
\end{equation}
%
\mpfive{This term contributes to the \emph{flux} of axonal branching as an
  additional source of diffusion, scaled by $\epsilon$, which reduces the
  branching density for a given projection where the branches of other
  projections are dense. Note that this operation is local to individual
  afferent projections.}

\MPone{In addition, the model we have outlined requires that molecular
  guidance gradients in the cortex are complemented by graded values of the
  interaction strengths, $\gamma$, at the level of the thalamus. While the
  precise mechanisms by which thalamic and cortical gradients interact during
  development have not been fully characterised, the presence of complementary
  thalamic and cortical guidance gradients has been well established
  experimentally. In particular, the EphA4 receptor and its ligand ephrin-A5
  are distributed in complementary gradients in the somatosensory thalamus and
  cortex} (\citealp{vanderhaeghen_mapping_2000,miller_epha7-ephrin-a5_2006}).

\MPone{Cells originating in VPM express high levels of EphA receptors and
  project to the lateral part of S1, which expresses low levels of ephrin-A5,
  and cells originating in the VPL express low levels of EphA receptors and
  project to the medial part of S1, which expresses high levels of ephrin-A5}
(see
\citealp{gao_regulation_1998,dufour_area_2003,vanderhaeghen_developmental_2004,speer_grading_2005,torii_role_2013}).
\MPone{We assume that such patterning arises because the relative strengths of
  interaction with guidance molecules (e.g., ephrin-A5) in the cortex are
  correlated with the relative concentrations of complementary molecules
  (e.g., EphA4) in the thalamus, and thus with thalamic position along the
  axis to which their gradients are aligned.}

@@

\MPthreePar{The parameters of the example simulation are indicated in
  Fig.\,\ref{fig:paramsweep}A using an asterisk. In
  Fig.\,\ref{fig:paramsweep}B, we also present examples of alternative
  patterns that emerge for different choices of $D$. Decreasing the rate of
  diffusion may be considered equivalent to increasing the overall size of the
  domain, $S$. Hence, insights into barrel development in species with a
  larger representation of the vibrissae, which do not have barrel fields, may
  be gained by studying pattern formation
  when $D$ is small. In this context, it is interesting to note that for small
  $D$, the organization is predicted to be topological but highly irregular,
  with a general expansion in the territory occupied by the central versus
  peripheral domains that would presumably manifest as an absence of
  identifiable barrel fields (Fig.\,\ref{fig:paramsweep}B\,iii).}

@@

\cmnt{In reality, scores of smaller Dirichletform barrels representing the
  microvibrissae form alongside the E-row barrels, presumably via the same
  competitive processes. Enforcing here
  the same boundary condition as used to represent the true edges of the
  barrel field was necessary to ensure the stability of the simulation, though
  we acknowledge that this region of the boundary was enforced primarily to
  keep the number of simulated projections, and hence the overall
  computational complexity of the simulation, manageable (simulating an extra
  projection introduces 13030 new dynamical variables).}
